---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r global_options,include=FALSE}
 knitr::opts_chunk$set(fig.path='README_files/figure-gfm/')
```
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)



# jtdm - Joint trait distribution modeling

## About the method

The package jtdm implements the method described in Poggiato et al., in prep. Joint models and predictions of community traits. The code for producing the results in this paper is available under the subfolder publications in this repo.

## Installing the R package

### R-package
The package implements jtdm using the Markov Chain Monte Carlo Bayesian modeling software JAGS via the R package runjags. Therefore, it requires the installation of JAGS. Its installation is easy and depends on your operating system:


##### Ubutntu
sudo apt-get install jags


##### Windows
https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/Windows/


##### Mac
https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/Mac%20OS%20X/

Once JAGS has been installed, the following code should run:

```{r,eval=FALSE}
library(devtools)
install_github("giopogg/jtdm")
```

The package implements jtdm using the Markov Chain Monte Carlo Bayesian modeling software JAGS via the R package runjags. Therefore, it requires the installation of both JAGS and runjags. 

Fitting a JTDM the case study dataset of Poggiato et al. In prep.
```{r,warning=FALSE}
library(jtdm)
library(ggplot2)
set.seed(1712)
data(Y)
data(X)
# Short MCMC to obtain a fast example: results are unreliable !
m = jtdm_fit(Y=Y, X=X, formula=as.formula("~GDD+FDD+forest"),  adapt = 10, burnin = 100, sample = 100)

# Inferred parameters
getB(m)$Bmean
get_sigma(m)$Smean
```
Trait-environment relationships
``` r
 partial_response(m,indexGradient="GDD",indexTrait="SLA",FixX=list(GDD=NULL,FDD=NULL,forest=1))$p
 ```
 
Partial response curve of the most suitable community-level strategy and envelop of possible community-level strategies of SLA and LNC along the GDD gradient.
```{r}
ellipse_plot(m,indexTrait = c("SLA","LNC"),indexGradient="GDD")
```

Computes joint probabilities of both SLA and LNC to be greater than 20 in a high altitude site. This measures the relative suitability of communities where both SLA and LNC are higher than 20 in a high altitude site
```{r}
joint_trait_prob(m,indexTrait=c("SLA","LNC"), Xnew=X["VCHA_2940",], bounds=list(c(20,Inf),c(20,Inf)))$PROBmean

```
Unsurprisingly, the probability is low. Then, we compute this probability along the GDD gradient
``` {r}
joint=joint_trait_prob_gradient(m,indexTrait=c("SLA","LNC"), indexGradient="GDD", bounds=list(c(mean(Y[,"SLA"]),Inf),c(mean(Y[,"SLA"]),Inf)))
```
And plot it

```{r,echo=FALSE}
table = data.frame(x=joint$gradient, mean= joint$GradProbsmean,
                                   q02 = joint$GradProbsq025,
                                   q97 = joint$GradProbsq975)
ggplot(data=table) + geom_ribbon(mapping=aes(x=x, ymin=q02, ymax=q97),position = position_dodge(0.3), size=1,alpha=0.2) + geom_line(mapping=aes(x=x, y=mean), size=1, position=position_dodge(width=0.3),col="#F8766D") + xlab("GDD")  + theme_classic() +              ggtitle("Joint probability of SLA and LNC to be both greater than 10 as a function of GDD")
```
As climatic conditions become more favourable (i.e. GDD increases), the probability of having high values of both traits increases.
