---
title: "casestudyPoggiatoInPrep"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r global_options,include=FALSE}
 knitr::opts_chunk$set(fig.path='README_files/figure-gfm/')
```

# Case study Poggiato et al. In prep.

In this R markdown we reproduce the case study of the publication Poggiato et al. In prep.  "Joint models and predictions of community traits". For the presentation of the R package \code{jtdm} and its details see https://github.com/giopogg/jtdm/tree/main/vignette .

# Installing the R package

### R-package
The package implements JTDMs using the Markov Chain Monte Carlo (MCMC) Bayesian modeling software JAGS via the R package runjags. Therefore, it requires the installation of JAGS. Its installation is easy and depends on your operating system:


##### Ubutntu
sudo apt-get install jags


##### Windows
https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/Windows/


##### Mac
https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/Mac%20OS%20X/

Once JAGS has been installed, the following code should run:

```{r,warning=FALSE,message=FALSE,results='hide',eval=TRUE}
library(devtools)
install_github("giopogg/jtdm")
library(jtdm)
library(ggplot2)
library(coda)
library(gridExtra)
library(ggpubr)
library(parallel) #if this doesn't work, set the parameter parallel = FALSE when calling joint_trait_prob
library(raster)
install_github("matthewkling/colormap")
library(colormap)
library(tidyr)
```

Load the dataset and run the model

```{r, warning=FALSE,message=FALSE,results='hide'}
data(X) #env variables
data(Y) #site x CWM traits matrix

formula=as.formula("~poly(GDD,2)+poly(FDD,2)+poly(GDD,2):forest+poly(FDD,2):forest")


# Run the model. If you want this vignette to run fast (buth with unreliable results), you should diminish the parameter 'sample'. To obtain the results of the publications, we set adapt = 5000, burnin = 10000, sample = 10000.
m = jtdm_fit(Y=Y, X=X, 
               formula=formula,
               adapt = 5000,
               burnin = 10000,
               sample = 10000,
               n.chains=2,
               monitor = c('B','Sigma','deviance'))
```


Check for convergence of the MCMC chain (Figure A2 in the publication)

```{r, warning=FALSE}
table=data.frame(x=gelman.diag(m$model$mcmc,multivariate = FALSE)$psrf[,1])

ggplot(data = table, aes(x=x)) + geom_histogram(color="darkblue", fill="lightblue",bins = 20) +
  ggtitle("Potential scale reduction factor") + xlab("") + ylab("nESS") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5))

```

Obtain $R2$ of the predictions both in-sample and in cross validation (values of Table A1 in the publication)

``` {r}
prediction = jtdm_predict(m=m, Xnew=X, Ynew= Y, validation = T, FullPost = F)

#R2 of in-sample predictions
prediction$R2
```

```{r, include=FALSE}
CV = jtdmCV(m,K=5, adapt = 5000,burnin = 10000, sample = 10000)
```

```{r}
#R2 of 5-fold cross validation
CV$R2
```


Compute the regression coefficients and plot effect sizes (Figure A3)
``` {r}
# get the regression coefficient matrix
B = getB(m  = m)

# compute standardised effect sizez
B_stand = B
for(i in 1:nrow(B$Bsamples)){
  for(j in 2:ncol(B$Bsamples))
    B_stand$Bsamples[i,j,] = sd(m$X[,j])*B$Bsamples[i,j,]/sd(m$Y[,i])
}
B_stand$Bsamples=B_stand$Bsamples[,-1,]
B_stand$Bmean = apply( B_stand$Bsamples, mean, MARGIN=c(1,2))
B_stand$Bq975 = apply( B_stand$Bsamples, quantile, MARGIN=c(1,2),0.975)
B_stand$Bq025 = apply( B_stand$Bsamples, quantile, MARGIN=c(1,2),0.025)

# build the table for ggplot
tableB_stand = data.frame(B= as.vector(B_stand$Bmean),
                          B97=as.vector(B_stand$Bq975), 
                          B02 = as.vector(B_stand$Bq025),
                          trait = rep(colnames(Y),ncol(m$X)-1),
                          predictor = rep(c("GDD","GDD","FDD","FDD","GDD","GDD","FDD","FDD"),
                                          each=ncol(Y)),
                          type= rep(c(1,2,1,2,3,4,3,4),each=ncol(Y)),
                          interaction_with_forest=rep(c("no","no","no","no","yes",
                                                      "yes","yes","yes"),
                                                      each=ncol(Y))
)

#check if significant or not
tableB_stand[,"significant"] = ifelse(sign(tableB_stand$B97)==sign((tableB_stand$B02)),"yes","no")


#plot
ggplot(data = tableB_stand, 
       aes(x = B, y = type, color = significant)) +
  geom_point(aes(shape=interaction_with_forest),size=2) +
  geom_errorbarh(aes(xmax = B97, xmin = B02, height = 0)) +
  geom_vline(xintercept=0,linetype="dashed") +
  facet_grid(trait ~ predictor) +
  ylim(c(0,4)) + theme_minimal()+ 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```


Compute and plot the residual covariance matrix (Figure A4).
``` {r}
Sigma = get_sigma(m = m)

Sigma_sign = ifelse(sign(Sigma$Sq025)==sign(Sigma$Sq975),1,0)
Sigma_plot = cov2cor(Sigma$Smean) * Sigma_sign
colnames(Sigma_plot) = rownames(Sigma_plot) = colnames(Y)

corrplot::corrplot(Sigma_plot,method="color" ,type="upper", order="hclust",
                   addCoef.col = "black", # Ajout du coefficient de corrélation
                    tl.col="black",tl.cex=2, diag=FALSE)
```



Plot trait environmental relationships (Figure A5).

```{r}
grid.length=200 #length of the gradient of the focal environmental variable
Xnameplot = c("GDD","FDD")
k=0 #counter

for(i in 1:(ncol(X)-1)){#for each environmental variable
  indexGradient=i

  ###### Build the XGradient_new matrices (a dataset with the gradient of the focal variable
  #     and all other ones set to their respective mean), one for open habitat and one for forests.

  
  # First build the gradient of the vocal variable
  XGradientFocal_open=  seq(from=min(X[,indexGradient]),
                            to=max(X[,indexGradient]),
                            length.out=grid.length)
  XGradientFocal_for=  seq(from=min(X[which(X[,"forest"]==1),indexGradient]),
                           to=max(X[,indexGradient]),
                           length.out=grid.length) 

  # Fill the XGradient_new matrices
  XGradient_new_open = matrix(nrow=grid.length,ncol=ncol(X))
  XGradient_new_for = matrix(nrow=grid.length,ncol=ncol(X))

  for(j in 1:ncol(X)){

    if(j == indexGradient){

      XGradient_new_open[,j] = XGradientFocal_open
      XGradient_new_for[,j] = XGradientFocal_for

    }else{
      #in open habitat, forest=0
      XGradient_new_open[,j] = mean(X[which(X[,"forest"]==0),j])
      #in forest, forest=1
      XGradient_new_for[,j] = mean(X[which(X[,"forest"]==1),j]) 
    }

  }

  colnames(XGradient_new_open) =  colnames(XGradient_new_for) = colnames(X)

  # Predict the value of traits when evaluated in XGradient_new matrices
  PartialPredictions_for = jtdm_predict(m=m,Xnew=XGradient_new_for, FullPost="mean")
  PartialPredictions_open = jtdm_predict(m=m,Xnew=XGradient_new_open, FullPost="mean")

  # Build a plot for each trait-environment combination
  for(j in 1:ncol(Y)){
    k=k+1
    assign(paste0("table_",k),
           data.frame(x = c(XGradientFocal_for,XGradientFocal_open), 
                      Predmean = c(PartialPredictions_for$PredMean[,j], 
                                   PartialPredictions_open$PredMean[,j]),
                      Pred975 = c(PartialPredictions_for$Predq975[,j],PartialPredictions_open$Predq975[,j]),
                      Pred025=c(PartialPredictions_for$Predq025[,j],PartialPredictions_open$Predq025[,j]),
                      type = c(rep("forest",times=grid.length),rep("open",times=grid.length))))

    assign(paste0("p_",k), 
           ggplot() + 
             geom_line(data=get(paste0("table_",k)), aes(x=x, y=Predmean,col=type)) +
             geom_ribbon(data=get(paste0("table_",k)),
                         aes(x=x,y=Predmean,ymin=Pred025,ymax=Pred975,col=type),linetype=2,alpha=0.3)+
             geom_rug(data=data.frame(x=X[,indexGradient]),aes(x=x),sides="b") +
             xlab(Xnameplot[indexGradient]) + ylab(colnames(Y)[j]) + theme_minimal() #+  theme(legend.position="none")
    )
  }

}

# Put all the plots together
eval(parse(text=paste0("p=as_ggplot(arrangeGrob(",paste(paste0("p_",c(1,4,2,5,3,6)),collapse=","),",nrow=3,ncol=2))")))
p

```


Plot the partial response curves of the most suitable CLS and of envelope of CLSs for each pair of traits and each environmental variable (Figure 4a,5a, A5).

```{r}
for(t in 1:(ncol(X)-1)){ #for each env covariate
  for(i in 1:(ncol(Y)-1)){
    for(j in (i+1):ncol(Y)){ #for each pairwise combination of traits

    # plot the curve
    print(ellipse_plot(m,indexGradient=colnames(m$X_raw)[t],
                 indexTrait = colnames(m$Y)[c(i,j)],FullPost=F,
                 FixX=list(GDD=NULL,FDD=NULL,forest=0)))

    }
  }
}

```

Build partial response plots for joint probabilities along all gradients, for all pairwise probabilities and for each of the four regions (joint-high, joint-low, disjoint). This corresponds to figures 4b, 5b, A6.

```{r, warnings=FALSE}

# This loop is quite long, it might take more than 1h depending on how long are the MCMC chains and whether you set parallel=T or not. In order for this to run faster, you can specify a lower number of mcmc.samples using the parameter mcmc.samples in the function joint_trait_prob_gradient.

# For each environmental variable (except habitat that is a dummy variable)
for(s in 1:(ncol(X)-1)){

# i and j are all pairwise trait combinations
 for(i in 1:(ncol(Y)-1)){

   for(j in (i+1):ncol(Y)){

     # k and l determine the region. k is for the first trait (i), l for the second one (j). P is when the         # given trait is above its mean, N when it is below  e.g. k="P" and l="P" is the joint high region.
     # The exact bounds of the regions are defined below.
     
     # prepare dataset for plotting
      JointCo_occGrad =as.data.frame(matrix(NA,ncol=1,nrow=100))

     
     EmpiricalCoOccPlot = data.frame(Coocc=NA, 
                                     X=rep(X[,s],times=4),
                                     id=rep("Observed points"),
                                     region=c(rep("PP", times=length(X[,s])),
                                              rep("PN", times=length(X[,s])),
                                              rep("NP", times=length(X[,s])),
                                              rep("NN", times=length(X[,s])))
                                     )

     for(k in c("P","N")){
       for(l in c("P","N")){
         #print(c(i,j,k,l)) #uncomment if you want to see the code updating

         # Define bounds of the region
         if(k=="P"){b1=c(mean(Y[,i]) , Inf)}else{b1=c(-Inf,mean(Y[,i]))}
         if(l=="P"){b2=c(mean(Y[,j]) , Inf)}else{b2=c(-Inf,mean(Y[,j]))}
         bounds=list(b1,b2) #bounds define the region in the community-trait space

         #run the function
         JointCo_occG = joint_trait_prob_gradient(m, indexTrait=colnames(m$Y)[c(i,j)],
                                                     indexGradient=colnames(m$X_raw)[s], 
                                                     bounds=bounds, grid.length=100,
                                                     FixX = list(GDD=NULL,FDD=NULL,forest=0),
                                                     FullPost=T,parallel = TRUE)
          
         JointCo_occGrad[,c(paste0(k,l,".mean"),
                            paste0(k,l,".975"),
                            paste0(k,l,".025"))] = cbind(JointCo_occG$GradProbsmean,
                                                         JointCo_occG$GradProbsq975,
                                                         JointCo_occG$GradProbsq025)


         tableGrad = data.frame(x=JointCo_occGrad$gradient, 
                                mean= JointCo_occGrad$GradProbsmean,
                                   q02 = JointCo_occGrad$GradProbsq025,
                                   q97 = JointCo_occGrad$GradProbsq975)


          # Create the 0/1 dataset (1 if both traits are >0, 0 otherwise), 
          # to change depending on the bounds

          for(r in 1:nrow(Y)){
              if(k=="P" & l=="P"){
              if(Y[r,i]>mean(Y[,i]) & Y[r,j]>mean(Y[,j])){
                EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="PP"),"Coocc"][r]=1
                }else{EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="PP"),"Coocc"][r]=0}
              }
              if(k=="P" & l=="N"){
                if(Y[r,i]>mean(Y[,i]) & Y[r,j]<mean(Y[,j])){
                  EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="PN"),"Coocc"][r]=1
                }else{EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="PN"),"Coocc"][r]=0
                  }
              }
              if(k=="N" & l=="P"){
                if(Y[r,i]>mean(Y[,i]) & Y[r,j]<mean(Y[,j])){
                  EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="NP"),"Coocc"][r]=1
                }else{EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="NP"),"Coocc"][r]=0
                  }
              }
              if(k=="N" & l=="N"){
                
                if(Y[r,i]<mean(Y[,i]) & Y[r,j]<mean(Y[,j])) {
                  EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="NN"),"Coocc"][r]=1
                  }else{EmpiricalCoOccPlot[which(EmpiricalCoOccPlot$region=="NN"),"Coocc"][r]=0}
              }

              }

       }
     }
     
     # Formatting the table to plot
     JointCo_occGrad = JointCo_occGrad[,-1]
     JointCo_occGrad[,"X"] =  JointCo_occG$gradient # the gradient is the same
      
     
     tableGrad = tidyr::gather(JointCo_occGrad,type,
                               Probability, colnames(JointCo_occGrad)[-which(colnames(JointCo_occGrad)=="X")])
     
     tableGrad[,"region"]=gsub("\\..*", "",tableGrad$type) #removes everything after point

     tableGrad[,"type"]=gsub("^.*\\.","",tableGrad$type)
     
     tableGrad = tidyr::spread(tableGrad, key="type", value = "Probability")
     colnames(tableGrad)[c(3,4)]=c("q025","q975")
     
     tableGrad$region=as.factor(tableGrad$region)
     levels(tableGrad$region)=c("NP","PP","NN","PN")
     
     EmpiricalCoOccPlot$region=as.factor(EmpiricalCoOccPlot$region)
     levels(EmpiricalCoOccPlot$region)=c("NP","PP","NN","PN")

     print(ggplot(data=tableGrad) + 
              geom_ribbon(mapping=aes(x=X, ymin=q025, ymax=q975),
                          position = position_dodge(0.3), size=1,alpha=0.2) +
              geom_line(mapping=aes(x=X, y=mean), size=1,
                        position=position_dodge(width=0.3),
                        col="#F8766D")+
              geom_point(data=EmpiricalCoOccPlot,
                         mapping=aes(x=X, y=Coocc), 
                         alpha=0.2,col="#00BFC4")+
              ggtitle(paste0("Joint probabilities of ",
                             colnames(Y)[i]," and ",colnames(Y)[j] ,
                             " as a function of ",colnames(X)[s])) +
              xlab("") + ylab("") + theme(plot.title = element_text(hjust = 0.5),
                                          plot.background = element_rect(fill = "white",
                                                                         colour = NA),
                                          panel.background = element_rect(fill = "white",
                                                                          colour = NA),
                                          panel.grid.major = element_line(colour="grey",
                                                                          size=0.5)) +
              facet_wrap(.~region) )
   }
 }
}


```



We now load the environmental rasters in the French Alps in order to predict the joint distribution of traits.

```{r}
load(file="env_alp_stack.Rdata")

# plot
plot(env_alp_stack)

# Some data formatting
env_alp_stack.df=as.data.frame(env_alp_stack,xy=T)

# add Id column
env_alp_stack.df[,"Id"]=1:nrow(env_alp_stack.df)
# add isna column (pixels to be removed)
env_alp_stack.df[,"isna"] = apply(env_alp_stack.df,MARGIN=1,FUN=function(x) ifelse(length(which(is.na(x)))>0,TRUE,FALSE))

# remove na pixels
X.df.xy = env_alp_stack.df[which(!env_alp_stack.df$isna),]
X.df = X.df.xy

#change rownames
rownames(X.df) = X.df$Id
#take only covariates to predict
X.df=subset(X.df,select=-c(Id,isna))
# reorder according to the original X
X.df=X.df[,colnames(X)]
# transform to numeric
X.df=apply(X.df,MARGIN=c(1,2),FUN=as.numeric)

```


Predict the marginal value of each trait.
```{r}
# We only take the posterior mean
AlpPred=as.data.frame(jtdm_predict(m=m, Xnew=X.df, Ynew = NULL, validation = F, FullPost=F)$PredMean)
head(AlpPred)

AlpPred[,"Id"]=rownames(AlpPred)
#merge
AlpPredXY=merge(X.df.xy,AlpPred,by="Id")

p1=ggplot(AlpPredXY, aes(x=x,y=y)) +
  geom_raster(aes(fill=SLA)) +ggtitle("SLA")+theme_classic()
#Height
p2=ggplot(AlpPredXY, aes(x=x,y=y)) +
  geom_raster(aes(fill=Height)) +ggtitle("Height") +theme_classic()

#LNC
p3=ggplot(AlpPredXY, aes(x=x,y=y)) +
  geom_raster(aes(fill=LNC)) +ggtitle("LNC")+theme_classic()

grid.arrange(p1,p2,p3,nrow=2,ncol=2)
```



Plot the predictions of the three traits in the RGB space (Figure A7)
``` {r, warnings=FALSE, message=FALSE}
AlpPredXY.col=data.frame(AlpPredXY,col=colors3d(AlpPredXY[,c("SLA","LNC","Height")]))

ggplot(AlpPredXY.col, aes(x=x,y=y)) + theme_classic()+
      geom_raster(aes(fill=as.factor(col))) +
      theme(legend.position="none")  +
      eval(parse(text=paste0("scale_fill_manual(values=c(",paste(paste0('"',unique(AlpPredXY.col$col),'"',"=",'"',unique(AlpPredXY.col$col),'"'),collapse = ","),"))")))

```



Then compute and plot the joint probabilities (Figure 6, A8). This can take ~15minutes.
```{r,warnings=FALSE}
env_alp_stack_pred.df=merge(env_alp_stack.df,subset(AlpPredXY.col,select=-c(x,y,forest,GDD,FDD,isna)),by="Id",all.x=T)

Xnew=X.df

JointCo_occProb=as.data.frame(matrix(NA,ncol=1,nrow=nrow(Xnew)))
rownames(JointCo_occProb)=rownames(Xnew)
for(i in 1:(ncol(Y)-1)){
  for(j in (i+1):ncol(Y)){

    for(k in c("P","N")){
      for(l in c("P","N")){

        #print(paste0(i,j,k,l)) # uncomment to see the code updating

        # Define bounds of the region
        if(k=="P"){b1=c(mean(Y[,i]) , Inf)}else{b1=c(-Inf,mean(Y[,i]))}
        if(l=="P"){b2=c(mean(Y[,j]) , Inf)}else{b2=c(-Inf,mean(Y[,j]))}
        bounds=list(b1,b2) #bounds define the region in the community-trait space

        JointCo_occProb[,paste0("JointProb_",
                                colnames(Y)[i],
                                "_",colnames(Y)[j],
                                "_",k,l)] = joint_trait_prob(m,
                                                             indexTrait=colnames(m$Y)[c(i,j)],
                                                             Xnew=Xnew, 
                                                             bounds=bounds,
                                                             FullPost = F)$PROBmean

      }
    }
  }
}

JointCo_occProb=JointCo_occProb[,-1]
JointCo_occProb[,"Id"]=rownames(JointCo_occProb)

env_alp_stack_pred.df=merge(env_alp_stack_pred.df,JointCo_occProb,by="Id",all.x=T)

for(i in 1:(ncol(Y)-1)){
  for(j in (i+1):ncol(Y)){


    t =  env_alp_stack_pred.df[,c("x","y",
                                  paste0("JointProb_",
                                         colnames(Y)[i],"_",
                                         colnames(Y)[j],"_PP"),
                                  paste0("JointProb_",
                                         colnames(Y)[i],"_",
                                         colnames(Y)[j],"_PN"),
                                  paste0("JointProb_",
                                         colnames(Y)[i],"_",
                                         colnames(Y)[j],"_NP"),
                                  paste0("JointProb_",
                                         colnames(Y)[i],"_",
                                         colnames(Y)[j],"_NN"))]
    
    colnames(t)[3:ncol(t)]=c("2.PP","4.PN","1.NP","3.NN")
    
    t=tidyr::gather(t,type,Probability,c("2.PP","4.PN","1.NP","3.NN"))


    print(ggplot(t, aes(x=x,y=y)) + theme_classic()+
            geom_raster(aes(fill=Probability)) +
            scale_fill_viridis_c(na.value = "white")+ facet_wrap(.~type) +
            ggtitle(paste0("Joint Probabilities of ",
                           colnames(Y)[i], " and ", 
                           colnames(Y)[j])) +
            theme_classic() )

  }
}

```
